Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/method.tex
fixed the merge part, and added details to the list of goals

The following were the objectives of the research:
\begin{itemize}
\item Implement heapsort, mergesort, quicksort, radixsort, shellsort and the four
elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
\item Simulate these sorts using the SimpleScalar architecture simulator, using a
variety of branch predictors and cache configurations.
\item Implement cache-conscious sorts from \cite{LaMarca96} and
\cite{LaMarca99}.
\item Investigate and analyse the increase in speed, reduction in cache misses, 
and change in branch prediction accuracy as a result of these changes.
\item Determine if any changes can be made to improved the sorts' performance.
\end{itemize}

@@

A tiled mergesort was then written, based on LaMarca's guidance. Firstly, this
was aligned in memory so that the number of misses due to conflicts would be
reduced. Then it fully sorted segments of the array, each half the size of the
cache, so that temporal locality would be exploited. Finally, all these segments
were merged together in the remaining merge steps.

@@

%TODO this bit also has to be checked once we fix mergesort. if w fix mergesort
Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
This attempted to completely sort in the level 1 cache, then to completely sort
in the level 2 cache. Double-aligned multi-mergesort was a combination of this
and of multi-mergesort. The rewrite was based on the base mergesort, and was
written far more cleanly, being more readable and efficient. These were also
found to reduce the instruction count, as well as the level 2 cache miss count.
These results are discussed in more detail in Section \ref{Mergesort results}.

@@

\section{Testing details}

>>>>

The following were the objectives of the research:
\begin{itemize}
\item Implement heapsort, mergesort, quicksort, radixsort, shellsort and the four
elementary sorts: insertion sort, selection sort, bubblesort and shakersort.
\item Implement cache-conscious sorts from \cite{LaMarca96} and
\cite{LaMarca99}.
\item Simulate these sorts using the SimpleScalar architecture simulator, using a
variety of branch predictors and cache configurations.
\item Measure the performance of these sorts using the PapiEx performance
counters.
\item Simulate bimodal branch predictors in software and use them to measure the
sorts.
\item Investigate and analyse the increase in speed, reduction in cache misses, 
and change in branch prediction accuracy as a result of the changes made.
\item Determine if any changes can be made to improved the sorts' performance.
\end{itemize}

@@

A tiled mergesort was then written, based on LaMarca's guidance. Firstly, this
was aligned in memory so that the number of misses due to conflicts would be
reduced. Then it fully sorted segments of the array, each half the size of the
cache, so that temporal locality would be exploited. It also fully sorted
segements within the level 1 cache. Finally, all these segments were merged
together in the remaining merge steps.

@@

%TODO this bit also has to be checked once we fix mergesort. if w fix mergesort
Once these sorts were written it was attempted to improve them slightly.
Instead of a tiled mergesort, a double-aligned tiled mergesort was written.
This aligned the arrays to avoid conflicts in the level 1 cache, at a slight
cost to level 2 cache performance. Double-aligned multi-mergesort was a
combination of this and of multi-mergesort. The rewrite was based on the base
mergesort, and was written far more cleanly, being more readable and efficient.
These were also found to reduce the instruction count, as well as the level 2
cache miss count. These results are discussed in more detail in Section
\ref{Mergesort results}.

@@

\section{Testing Details}
